---
title: "eGFR workflow"
author: "Shen Jia"
date: "2026-07-01"
output:
  html_document:
    toc: true
    toc_depth: 2
---
Install the packages needed in the research. Define the functions included in the research
```{r setup, message=FALSE, warning=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)

# Change only these paths when moving the analysis.
data_path <- "/Users/shenjia/Downloads/harmonized data with biopsy.csv"
output_dir <- file.path(getwd(), "egfr_output")
dir.create(output_dir, recursive = TRUE, showWarnings = FALSE)

required_packages <- c(
  "dplyr", "tidyr", "ggplot2", "nlme", "lspline",
  "stringr", "kml", "cmprsk", "scales", "patchwork"
)
missing_packages <- required_packages[
  !vapply(required_packages, requireNamespace, logical(1), quietly = TRUE)
]
if (length(missing_packages)) {
  stop("Install missing packages: ", paste(missing_packages, collapse = ", "))
}
invisible(lapply(required_packages, library, character.only = TRUE))

group_cols <- c("Recovered group" = "#0072B2", "Static group" = "#D55E00")
induction_cols <- c(
  "Cyclophosphamide" = "#0072B2",
  "Rituximab" = "#D55E00",
  "Other" = "#009E73"
)
berden_levels <- c("Focal", "Crescentic", "Mixed", "Sclerotic", "Missing")
berden_cols <- c(
  "Focal" = "#1a9850", "Crescentic" = "#66bd63",
  "Mixed" = "#fdae61", "Sclerotic" = "#d73027", "Missing" = "grey70"
)

num_clean <- function(x) suppressWarnings(as.numeric(as.character(x)))
date_clean <- function(x) suppressWarnings(as.Date(as.character(x)))

standardize_berden <- function(x) {
  x <- trimws(as.character(x))
  x[is.na(x) | x %in% c("", "NA", "N/A", "na", "n/a")] <- "Missing"
  x
}

checked_label <- function(x, label) {
  x <- as.character(x)
  x[x == "Checked"] <- label
  x[x == "Unchecked"] <- " "
  x
}

collapse_checked <- function(x) {
  x <- trimws(as.character(x))
  if (any(x == "Checked", na.rm = TRUE)) "Checked" else
    if (all(is.na(x) | x == "")) NA_character_ else "Unchecked"
}

first_nonblank <- function(x) {
  keep <- !is.na(x) & trimws(as.character(x)) != ""
  if (any(keep)) x[which(keep)[1]] else x[NA_integer_][1]
}
first_valid_date <- function(x) {
  x <- x[!is.na(x)]
  if (length(x)) x[1] else as.Date(NA_character_)
}

first_finite <- function(x) {
  x <- num_clean(x)
  x <- x[is.finite(x)]
  if (length(x)) x[1] else NA_real_
}

assert_columns <- function(data, columns, object_name = deparse(substitute(data))) {
  missing <- setdiff(columns, names(data))
  if (length(missing)) {
    stop(object_name, " is missing: ", paste(missing, collapse = ", "))
  }
  invisible(data)
}

save_plot <- function(plot, filename, width, height, dpi = 600) {
  path <- file.path(output_dir, filename)
  ggplot2::ggsave(path, plot, width = width, height = height, dpi = dpi, bg = "white")
  if (!file.exists(path) || is.na(file.info(path)$size) || file.info(path)$size <= 0) {
    stop("Figure was not saved correctly: ", path)
  }
  invisible(path)
}

save_table <- function(x, filename) {
  path <- file.path(output_dir, filename)
  write.csv(x, path, row.names = FALSE)
  invisible(path)
}

theme_manuscript <- function(base_size = 13, legend = "bottom") {
  theme_classic(base_size = base_size) +
    theme(
      axis.title = element_text(face = "bold"),
      axis.text = element_text(colour = "black"),
      panel.grid.major.y = element_line(colour = "grey90", linewidth = 0.35),
      panel.grid.major.x = element_blank(),
      panel.grid.minor = element_blank(),
      legend.position = legend
    )
}

format_p <- function(p, prefix = FALSE) {
  label <- ifelse(p < 0.001, "<0.001", sprintf("%.3f", p))
  if (prefix) paste("P", ifelse(p < 0.001, label, paste("=", label))) else label
}

wilcox_p <- function(x, group) {
  keep <- !is.na(x) & !is.na(group)
  x <- x[keep]
  group <- droplevels(factor(group[keep]))
  if (nlevels(group) != 2 || length(unique(x)) < 2) return(NA_real_)
  wilcox.test(x ~ group, exact = FALSE)$p.value
}

fisher_p <- function(x, group) {
  keep <- !is.na(x) & !is.na(group)
  tab <- table(group[keep], x[keep])
  if (nrow(tab) != 2 || ncol(tab) < 2) return(NA_real_)
  fisher.test(tab)$p.value
}

tidy_cuminc <- function(fit, event_code = 1L, population = NULL) {
  curve_names <- setdiff(names(fit), "Tests")
  curve_names <- curve_names[grepl(paste0("(^| )", event_code, "$"), curve_names)]
  if (!length(curve_names)) stop("No CIF curve found for event code ", event_code)

  bind_rows(lapply(curve_names, function(curve_name) {
    curve <- fit[[curve_name]]
    data.frame(
      name = curve_name,
      time = curve$time,
      estimate = curve$est,
      variance = curve$var
    )
  })) %>%
    mutate(
      population = population,
      group = sub(paste0(" ?", event_code, "$"), "", name),
      se = sqrt(pmax(variance, 0)),
      lower = pmax(estimate - 1.96 * se, 0),
      upper = pmin(estimate + 1.96 * se, 1)
    )
}

fine_gray_stats <- function(fit, term = 1L) {
  beta <- unname(fit$coef[term])
  se <- sqrt(fit$var[term, term])
  tibble(
    beta = beta,
    SHR = exp(beta),
    lower = exp(beta - 1.96 * se),
    upper = exp(beta + 1.96 * se),
    p = 2 * pnorm(-abs(beta / se))
  )
}

predict_fixed_lme <- function(model, newdata) {
  design <- model.matrix(delete.response(terms(model)), newdata)
  beta <- nlme::fixef(model)
  covariance <- vcov(model)
  if (!setequal(colnames(design), names(beta))) {
    stop("Prediction design matrix does not match the model coefficients.")
  }
  design <- design[, names(beta), drop = FALSE]
  estimate <- as.numeric(design %*% beta)
  se <- sqrt(rowSums((design %*% covariance) * design))
  critical <- qt(0.975, df = nobs(model) - length(beta))
  bind_cols(
    as_tibble(newdata),
    tibble(
      fit = estimate,
      lower = estimate - critical * se,
      upper = estimate + critical * se
    )
  )
}

extract_cif_times <- function(curve_data, times = c(1, 3, 5)) {
  bind_rows(lapply(unique(curve_data$population), function(population_value) {
    curve <- curve_data %>% filter(population == population_value) %>% arrange(time)
    bind_rows(lapply(times, function(target_time) {
      idx <- which(curve$time <= target_time)
      if (!length(idx)) {
        return(tibble(
          population = population_value, year = target_time,
          estimate = 0, lower = 0, upper = 0
        ))
      }
      row <- curve[max(idx), ]
      tibble(
        population = population_value, year = target_time,
        estimate = row$estimate, lower = row$lower, upper = row$upper
      )
    }))
  }))
}
```

## 1. Load, filter, and clean the data 
Select the variable of interest to the research. Full Cohort 

```{r load-filter-clean}
raw_data <- read.csv(data_path, stringsAsFactors = FALSE, check.names = TRUE)

disease_labels <- c(
  "Granulomatosis with polyangiitis (Wegener) - Orpha:900",
  "Microscopic polyangiitis (including renal limited vasculitis) - ORPHA:727"
)

exRKD <- raw_data %>%
  filter(
    diagnosis_confidence == "Definite",
    small_vessel_vas_anca %in% disease_labels,
    Secondary.vasculitis != "Yes",
    Other != "Yes",
    is.na(Medium.vessel.vasculitis) | Medium.vessel.vasculitis == "",
    is.na(Large.vessel.vasculitis) | Large.vessel.vasculitis == "",
    is.na(Variable.vessel.vasculitis) | Variable.vessel.vasculitis == "",
    is.na(small_vessel_vas_immune) |
      small_vessel_vas_immune %in% c("", "Anti-GBM disease - ORPHA:375")
  ) %>%
  mutate(
    inductiontreatmentType_Rituximab = checked_label(
      inductiontreatmentType_Rituximab, "Rituximab"
    ),
    inductiontreatmentType_Cyclophosphamide = checked_label(
      inductiontreatmentType_Cyclophosphamide, "Cyclo.P"
    ),
    inductiontreatmentType_.Daily.Oral.Cyclophosphamide = checked_label(
      inductiontreatmentType_.Daily.Oral.Cyclophosphamide, "Cyclo.O"
    ),
    inductiontreatmentType_.Pulsed.IV.corticosteroids = checked_label(
      inductiontreatmentType_.Pulsed.IV.corticosteroids, "IV.Steroids"
    ),
    inductiontreatmentType_Oral.corticosteroids = checked_label(
      inductiontreatmentType_Oral.corticosteroids, "Oral.Steroids"
    ),
    cyclo = if_else(
      inductiontreatmentType_.Daily.Oral.Cyclophosphamide == "Cyclo.O" |
        inductiontreatmentType_Cyclophosphamide == "Cyclo.P",
      "Cyclo", " "
    ),
    corti = if_else(
      inductiontreatmentType_Oral.corticosteroids == "Oral.Steroids" |
        inductiontreatmentType_.Pulsed.IV.corticosteroids == "IV.Steroids",
      "Corti", " "
    ),
    corti_seperate = str_c(
      inductiontreatmentType_Oral.corticosteroids, ";",
      inductiontreatmentType_.Pulsed.IV.corticosteroids
    ),
    induction_trtmnt = str_c(inductiontreatmentType_Rituximab, " ", cyclo),
    inductionFinal = trimws(induction_trtmnt),
    inductionFinal = if_else(
      inductionFinal %in% c("Cyclo", "Rituximab"), inductionFinal, "Other"
    ),
    Berden.score.on.renal.biopsy = standardize_berden(
      Berden.score.on.renal.biopsy
    ),
    ancaSpec = if_else(ancaSpec %in% c("MPO", "PR3"), ancaSpec, "Other")
  )

organ_cols <- c(
  "affectedOrgan_Renal", "affectedOrgan_Cutaneous", "affectedOrgan_General",
  "affectedOrgan_Abdominal", "affectedOrgan_Mucous_membranes_eyes",
  "affectedOrgan_Cardiovascular", "affectedOrgan_Other", "affectedOrgan_ENT",
  "affectedOrgan_Chest", "affectedOrgan_Nervous_system"
)
core_cols <- c(
  "RKD.ID", "creatininetest", "egfr_calculated", "dateOfDiagnosis",
  "Interval.from.diagnosis..months.", "Date.Of.Visit", "Dialysis.dependent",
  "Berden.score.on.renal.biopsy", "dateOfDeath", "dateOfESKD", "ancaSpec",
  "gender", "DefiniteGPA", "DefiniteMPA", "ESKDStandard",
  "lastRecordedContact", "death", "ageAtDiagnosis", "inductionFinal",
  "inductiontreatmentType_.Pulsed.IV.corticosteroids",
  "inductiontreatmentType_Oral.corticosteroids",
  "inductiontreatmentType_Plasma.exchange",
  "inductiontreatmentType_Methotrexate",
  "inductiontreatmentType_Rituximab", "cyclo"
)
needed_cols <- unique(c(core_cols, organ_cols))
assert_columns(exRKD, needed_cols)

renal_raw <- exRKD %>%
  select(all_of(needed_cols)) %>%
  mutate(
    across(c(creatininetest, egfr_calculated, Interval.from.diagnosis..months.), num_clean),
    across(c(Date.Of.Visit, dateOfDiagnosis, dateOfDeath, dateOfESKD,
             lastRecordedContact), date_clean),
    Dialysis.dependent = trimws(as.character(Dialysis.dependent)),
    Dialysis.dependent = if_else(
      is.na(Dialysis.dependent) | Dialysis.dependent == "", "No", Dialysis.dependent
    ),
    Diagnosis = case_when(
      DefiniteMPA == 1 ~ "MPA",
      DefiniteGPA == 1 ~ "GPA",
      TRUE ~ NA_character_
    )
  )

# Organ fields are summarized separately so that patient-level collapsing does
# not alter the visit-level renal filter used in the original ECT workflow.
organ_summary <- renal_raw %>%
  group_by(RKD.ID) %>%
  summarise(across(all_of(organ_cols), collapse_checked), .groups = "drop") %>%
  summarise(across(all_of(organ_cols), ~ sum(.x == "Checked", na.rm = TRUE))) %>%
  pivot_longer(everything(), names_to = "Organ", values_to = "n_checked")

cat("Original IDs:", n_distinct(raw_data$RKD.ID), "\n")
cat("Filtered AAV IDs:", n_distinct(renal_raw$RKD.ID), "\n")
print(organ_summary)
```

## 2. Derive Early renal recovery cohort (Cohort 1)

```{r cohort-derivation}
Full_cohort <- renal_raw

# Full-cohort baseline rule: earliest value in [-0.5, 1] months; if absent,
# use the earliest dated value anywhere. This intentionally differs from ECT.
baseline_values <- Full_cohort %>%
  group_by(RKD.ID) %>%
  group_modify(~ {
    time <- .x$Interval.from.diagnosis..months.
    preferred <- !is.na(time) & time >= -0.5 & time <= 1
    if (any(preferred)) {
      .x[preferred, ] %>% arrange(Interval.from.diagnosis..months.) %>% slice(1)
    } else if (any(!is.na(time))) {
      .x %>%
        filter(!is.na(Interval.from.diagnosis..months.)) %>%
        arrange(Interval.from.diagnosis..months.) %>%
        slice(1)
    } else {
      slice(.x, 1)
    }
  }) %>%
  ungroup() %>%
  select(
    RKD.ID,
    creatininetest1 = creatininetest,
    egfr_calculated1 = egfr_calculated
  )

Full_cohort <- Full_cohort %>% left_join(baseline_values, by = "RKD.ID")

# ECT source: renal visits in the first six months. Dialysis-dependent visits
# are assigned eGFR=3 and creatinine=500 before baseline and change calculation.
crtn <- renal_raw %>%
  filter(
    affectedOrgan_Renal == "Checked",
    !is.na(Date.Of.Visit),
    !is.na(Interval.from.diagnosis..months.),
    Interval.from.diagnosis..months. > -0.5,
    Interval.from.diagnosis..months. <= 6
  ) %>%
  mutate(
    egfr_calculated = if_else(Dialysis.dependent == "Yes", 3, egfr_calculated),
    creatininetest = if_else(Dialysis.dependent == "Yes", 500, creatininetest)
  ) %>%
  filter(is.finite(egfr_calculated)) %>%
  arrange(RKD.ID, Date.Of.Visit, Interval.from.diagnosis..months.) %>%
  distinct(RKD.ID, Date.Of.Visit, .keep_all = TRUE) %>%
  arrange(RKD.ID, Interval.from.diagnosis..months.)

# ECT baseline rule: earliest value at 0-1 month; otherwise the closest value
# before diagnosis in (-0.5, 0). Do not merge this with the full-cohort rule.
baseline_post <- crtn %>%
  filter(between(Interval.from.diagnosis..months., 0, 1)) %>%
  arrange(RKD.ID, Interval.from.diagnosis..months.) %>%
  distinct(RKD.ID, .keep_all = TRUE)

baseline_pre <- crtn %>%
  filter(
    !RKD.ID %in% baseline_post$RKD.ID,
    Interval.from.diagnosis..months. > -0.5,
    Interval.from.diagnosis..months. < 0
  ) %>%
  arrange(RKD.ID, desc(Interval.from.diagnosis..months.)) %>%
  distinct(RKD.ID, .keep_all = TRUE)

crtn_start <- bind_rows(baseline_post, baseline_pre) %>%
  distinct(RKD.ID, .keep_all = TRUE) %>%
  transmute(
    RKD.ID,
    creatininetest1 = creatininetest,
    eGFR1 = egfr_calculated
  )

baseline_eGFR_status <- Full_cohort %>%
  distinct(RKD.ID) %>%
  mutate(has_baseline_eGFR = RKD.ID %in% crtn_start$RKD.ID)

crtn <- crtn %>%
  inner_join(crtn_start, by = "RKD.ID") %>%
  mutate(
    creatininetest_delta = creatininetest - creatininetest1,
    creatininetest_percent = if_else(
      !is.na(creatininetest1) & creatininetest1 > 0,
      100 * creatininetest_delta / creatininetest1,
      NA_real_
    ),
    eGFR_delta = egfr_calculated - eGFR1,
    eGFR_percent = if_else(
      !is.na(eGFR1) & eGFR1 > 0, 100 * eGFR_delta / eGFR1, NA_real_
    )
  ) %>%
  filter(is.finite(eGFR_delta)) %>%
  group_by(RKD.ID) %>%
  mutate(
    n_of_eGFR = n(),
    has_post_diagnosis = any(Interval.from.diagnosis..months. >= 0, na.rm = TRUE)
  ) %>%
  ungroup() %>%
  filter(n_of_eGFR > 1, has_post_diagnosis)

ECT_cohort <- crtn
Ect_data <- ECT_cohort %>%
  mutate(
    gender = factor(gender),
    ancaSpec = factor(ancaSpec, levels = c("MPO", "PR3", "Other")),
    inductionFinal = factor(inductionFinal, levels = c("Cyclo", "Rituximab", "Other")),
    Diagnosis = factor(Diagnosis, levels = c("MPA", "GPA"))
  )

cat("Full cohort IDs:", n_distinct(Full_cohort$RKD.ID), "\n")
cat("Patients without baseline eGFR:", sum(!baseline_eGFR_status$has_baseline_eGFR), "\n")
cat("ECT cohort IDs/rows:", n_distinct(Ect_data$RKD.ID), "/", nrow(Ect_data), "\n")

stopifnot(
  n_distinct(Full_cohort$RKD.ID) == 742L,
  n_distinct(Ect_data$RKD.ID) == 311L
)
```

## 3. We do analysis to the Cohort 1.
We mainly interest in the change of eGFR in the first 6 month, and thus fit the mixed effects models with different covariates.

```{r ect-models}
fit_lme <- function(formula, data = Ect_data, na.action = na.fail) {
  model <- nlme::lme(
    formula, random = ~ 1 | RKD.ID, data = data,
    method = "REML", na.action = na.action
  )
  model
}

lm1 <- fit_lme(eGFR_delta ~ Interval.from.diagnosis..months.)
lm2 <- fit_lme(eGFR_delta ~ lspline(Interval.from.diagnosis..months., knot = c(2, 4)))
lm3 <- fit_lme(eGFR_delta ~ gender * lspline(Interval.from.diagnosis..months., knot = c(2, 4)))

Ect_data_2 <- Ect_data %>% filter(ancaSpec != "Other") %>% droplevels()
lm5 <- fit_lme(
  eGFR_delta ~ ancaSpec * lspline(Interval.from.diagnosis..months., knot = c(2, 4)),
  Ect_data_2
)
lm6 <- fit_lme(
  eGFR_delta ~ Diagnosis * lspline(Interval.from.diagnosis..months., knot = c(2, 4)),
  na.action = na.omit
)
lm7 <- fit_lme(
  eGFR_delta ~ inductionFinal * lspline(Interval.from.diagnosis..months., knot = c(2, 4))
)
lm8 <- fit_lme(
  eGFR_delta ~ ancaSpec + gender +
    inductionFinal * lspline(Interval.from.diagnosis..months., knot = c(2, 4)),
  na.action = na.omit
)

model_exports <- list(lm3 = lm3, lm5 = lm5, lm6 = lm6, lm7 = lm7, lm8 = lm8)
invisible(Map(
  function(model, model_name) {
    model_table <- summary(model)$tTable %>%
      as.data.frame() %>%
      tibble::rownames_to_column("term")
    save_table(model_table, paste0(model_name, "_tTable_egfr.csv"))
  },
  model_exports,
  names(model_exports)
))

print(summary(lm2)$tTable)
```

```{r ect-figures, fig.width=7.2, fig.height=8.5}
time_var <- "Interval.from.diagnosis..months."

overall_grid <- setNames(
  data.frame(seq(0, 6, length.out = 121)),
  time_var
)
eff_overall_df <- predict_fixed_lme(lm2, overall_grid)

# Extract the three spline slopes and P values directly from lm2.
lm2_table <- summary(lm2)$tTable
slope_rows <- grep("^lspline", rownames(lm2_table))
stopifnot(length(slope_rows) == 3L)
overall_stats <- tibble(
  x = c(1, 3, 5),
  interval = c("0-2 months", "2-4 months", "4-6 months"),
  slope = lm2_table[slope_rows, "Value"],
  p = lm2_table[slope_rows, "p-value"],
  label = paste0(
    interval, "\nSlope = ", sprintf("%.2f", slope), "/month\n",
    vapply(p, format_p, character(1), prefix = TRUE)
  )
)

p_overall <- ggplot(eff_overall_df, aes(x = .data[[time_var]], y = fit)) +
  geom_hline(yintercept = 0, linetype = "dashed", linewidth = 0.35, colour = "grey55") +
  geom_vline(xintercept = c(2, 4), linetype = "dotted", linewidth = 0.35, colour = "grey70") +
  geom_ribbon(aes(ymin = lower, ymax = upper), fill = "grey80", alpha = 0.55) +
  geom_line(linewidth = 1.35, colour = "#0072B2", lineend = "round") +
  geom_label(
    data = overall_stats,
    aes(x = x, y = Inf, label = label),
    inherit.aes = FALSE,
    vjust = 1.15,
    size = 2.8,
    lineheight = 0.95,
    label.padding = grid::unit(0.14, "lines"),
    label.r = grid::unit(0.08, "lines"),
    linewidth = 0.2,
    fill = scales::alpha("white", 0.92)
  ) +
  scale_x_continuous(breaks = 0:6, limits = c(0, 6), expand = expansion(mult = c(0.01, 0.02))) +
  scale_y_continuous(breaks = scales::pretty_breaks(6), expand = expansion(mult = c(0.08, 0.32))) +
  labs(x = "Months since diagnosis", y = "Change in eGFR from baseline (mL/min/1.73 m²)") +
  theme_manuscript()

induction_grid <- expand_grid(
  time_value = seq(0, 6, length.out = 121),
  inductionFinal = factor(
    c("Cyclo", "Rituximab", "Other"),
    levels = c("Cyclo", "Rituximab", "Other")
  )
) %>%
  rename(!!time_var := time_value)

eff_ind_df <- predict_fixed_lme(lm7, induction_grid) %>%
  mutate(
    induction_label = recode(
      as.character(inductionFinal),
      Cyclo = "Cyclophosphamide",
      Rituximab = "Rituximab",
      Other = "Other"
    ),
    induction_label = factor(
      induction_label,
      levels = c("Cyclophosphamide", "Rituximab", "Other")
    )
  )

p_induction <- ggplot(
  eff_ind_df,
  aes(
    x = .data[[time_var]], y = fit,
    colour = induction_label, fill = induction_label, group = induction_label
  )
) +
  geom_hline(yintercept = 0, linetype = "dashed", linewidth = 0.35, colour = "grey55") +
  geom_vline(xintercept = c(2, 4), linetype = "dotted", linewidth = 0.35, colour = "grey70") +
  geom_ribbon(aes(ymin = lower, ymax = upper), alpha = 0.12, colour = NA, show.legend = FALSE) +
  geom_line(linewidth = 1.25, lineend = "round") +
  scale_colour_manual(values = induction_cols, name = NULL) +
  scale_fill_manual(values = induction_cols, guide = "none") +
  scale_x_continuous(breaks = 0:6, limits = c(0, 6), expand = expansion(mult = c(0.01, 0.02))) +
  scale_y_continuous(breaks = scales::pretty_breaks(6), expand = expansion(mult = c(0.08, 0.08))) +
  labs(x = "Months since diagnosis", y = expression(Delta~eGFR~"(mL/min/1.73"~m^2*")")) +
  theme_manuscript()

Figure2_combined <- (p_overall / p_induction) +
  plot_layout(heights = c(1, 1.05), guides = "collect") +
  plot_annotation(tag_levels = "A") &
  theme(legend.position = "bottom", plot.tag = element_text(face = "bold", size = 14))

print(Figure2_combined)
save_plot(p_overall, "Figure_2_overall_eGFR_recovery.png", 6.8, 4.6)
save_plot(p_induction, "Figure_induction_treatment_eGFR_lm7.png", 6.8, 4.8)
save_plot(Figure2_combined, "Figure_2_combined.png", 7.2, 8.5)
ggsave(
  file.path(output_dir, "Figure_2_combined.pdf"),
  Figure2_combined, width = 7.2, height = 8.5, bg = "white"
)
```
Now follow the figure 1 flowchart, we move analysis to Cohort 2 trajectory clustering

```{r retrace-kml}
RETRACE_base <- crtn %>%
  mutate(
    dateOfCensored_days = as.numeric(lastRecordedContact - dateOfDiagnosis),
    daysfrmESKD_days = as.numeric(dateOfESKD - dateOfDiagnosis)
  ) %>%
  filter(!is.na(dateOfCensored_days), dateOfCensored_days > 180)

ESKD_EARLY <- RETRACE_base %>%
  filter(!is.na(daysfrmESKD_days), daysfrmESKD_days <= 180) %>%
  distinct(RKD.ID) %>%
  pull(RKD.ID)

kmlDataMean_select <- RETRACE_base %>%
  filter(!RKD.ID %in% ESKD_EARLY) %>%
  mutate(
    time = pmax(Interval.from.diagnosis..months., 0),
    time_bin = pmin(pmax(ceiling(time), 1), 6)
  ) %>%
  group_by(RKD.ID, time_bin) %>%
  summarise(mean = mean(eGFR_delta, na.rm = TRUE), .groups = "drop") %>%
  filter(is.finite(mean)) %>%
  group_by(RKD.ID) %>%
  filter(n_distinct(time_bin) > 1) %>%
  ungroup()

if (n_distinct(kmlDataMean_select$RKD.ID) < 2) {
  stop("KML cohort has fewer than two patients after filtering.")
}

month_cols <- paste0("month_", 1:6)
kml_wide <- kmlDataMean_select %>%
  mutate(time_bin = factor(time_bin, levels = 1:6, labels = month_cols)) %>%
  pivot_wider(id_cols = RKD.ID, names_from = time_bin, values_from = mean) %>%
  arrange(RKD.ID) %>%
  as.data.frame()

for (column in setdiff(month_cols, names(kml_wide))) kml_wide[[column]] <- NA_real_
kml_wide <- kml_wide[, c("RKD.ID", month_cols)]
id_order <- kml_wide$RKD.ID

kml_obj <- kml::cld(kml_wide, timeInData = 2:ncol(kml_wide))
set.seed(123)
kml::kml(kml_obj, nbClusters = 2:5, nbRedrawing = 20)
cl2 <- getClusters(kml_obj, 2)

cluster_df_raw <- tibble(RKD.ID = id_order, cluster_raw = as.integer(cl2))
latest_by_id <- kmlDataMean_select %>%
  group_by(RKD.ID) %>%
  slice_max(time_bin, n = 1, with_ties = FALSE) %>%
  ungroup() %>%
  left_join(cluster_df_raw, by = "RKD.ID")

# Name clusters by the mean change at each patient's final available month.
cluster_order <- latest_by_id %>%
  group_by(cluster_raw) %>%
  summarise(latest_mean_eGFR_delta = mean(mean, na.rm = TRUE), .groups = "drop") %>%
  arrange(desc(latest_mean_eGFR_delta)) %>%
  mutate(
    cluster = row_number(),
    cluster_name = c("Recovered group", "Static group")[cluster]
  )

cluster_df <- cluster_df_raw %>%
  left_join(cluster_order, by = "cluster_raw") %>%
  select(RKD.ID, cluster, cluster_raw, cluster_name)

cluster_count <- cluster_df %>%
  count(cluster, cluster_name, name = "n") %>%
  arrange(cluster) %>%
  mutate(
    prop = round(100 * n / sum(n), 1),
    cluster_label = cluster_name
  )

cluster_df <- cluster_df %>%
  left_join(
    cluster_count %>% select(cluster, cluster_name, n, prop, cluster_label),
    by = c("cluster", "cluster_name")
  )

Data_wide <- kmlDataMean_select %>%
  left_join(cluster_df, by = "RKD.ID") %>%
  mutate(cluster_label = factor(
    cluster_label,
    levels = c("Recovered group", "Static group")
  ))

cluster_means <- Data_wide %>%
  group_by(cluster, cluster_label, time_bin) %>%
  summarise(mean_value = mean(mean, na.rm = TRUE), .groups = "drop")

print(cluster_count)
stopifnot(
  n_distinct(RETRACE_base$RKD.ID) == 285L,
  length(ESKD_EARLY) == 34L,
  n_distinct(kmlDataMean_select$RKD.ID) == 234L,
  setequal(cluster_count$n, c(97L, 137L))
)
```

```{r cluster-selection-figure, fig.width=7.4, fig.height=4.6}
crit_mat <- tryCatch(
  as(kml_obj, "ListPartition")["criterionValuesAsMatrix"],
  error = function(e) kml_obj["criterionValuesAsMatrix"]
) %>%
  as.matrix()

if (nrow(crit_mat) == 20 && ncol(crit_mat) == 4) crit_mat <- t(crit_mat)
if (is.null(rownames(crit_mat)) ||
    any(is.na(suppressWarnings(as.numeric(rownames(crit_mat)))))) {
  rownames(crit_mat) <- 2:(nrow(crit_mat) + 1)
}

crit_long <- as.data.frame(crit_mat, check.names = FALSE) %>%
  mutate(k = as.integer(gsub("[^0-9]", "", rownames(crit_mat)))) %>%
  filter(k %in% 2:5) %>%
  pivot_longer(-k, names_to = "redrawing", values_to = "CH") %>%
  mutate(CH = num_clean(CH)) %>%
  filter(is.finite(CH)) %>%
  group_by(k) %>%
  arrange(desc(CH), .by_group = TRUE) %>%
  mutate(
    redrawing_rank = row_number(),
    solution = factor(paste0(k, "-group solution"), levels = paste0(2:5, "-group solution")),
    selected = factor(k == 2, levels = c(FALSE, TRUE))
  ) %>%
  ungroup()

ch_cols <- c(
  "2-group solution" = "#0072B2", "3-group solution" = "#D55E00",
  "4-group solution" = "#009E73", "5-group solution" = "grey55"
)

p_ch <- ggplot(
  crit_long,
  aes(redrawing_rank, CH, colour = solution, group = solution)
) +
  geom_line(aes(linewidth = selected, alpha = selected), lineend = "round") +
  geom_point(aes(size = selected, alpha = selected)) +
  scale_colour_manual(values = ch_cols, name = NULL) +
  scale_linewidth_manual(values = c("FALSE" = 0.7, "TRUE" = 1.35), guide = "none") +
  scale_size_manual(values = c("FALSE" = 1.8, "TRUE" = 2.5), guide = "none") +
  scale_alpha_manual(values = c("FALSE" = 0.65, "TRUE" = 1), guide = "none") +
  scale_x_continuous(breaks = c(1, 5, 10, 15, 20)) +
  scale_y_continuous(breaks = scales::pretty_breaks(6)) +
  labs(
    x = "Rank across repeated random initializations",
    y = "Calinski-Harabasz criterion"
  ) +
  theme_manuscript()

print(p_ch)
save_plot(p_ch, "Figure_4_cluster_selection_CH-2.png", 7.4, 4.6)
```

```{r trajectory-figure, fig.width=7.2, fig.height=5.2}
group_info <- Data_wide %>%
  distinct(RKD.ID, cluster_label) %>%
  count(cluster_label, name = "n") %>%
  mutate(
    pct = round(100 * n / sum(n), 1),
    legend_lab = paste0(cluster_label, " (n = ", n, ", ", pct, "%)")
  )
legend_labs <- setNames(group_info$legend_lab, group_info$cluster_label)

plot_c <- ggplot() +
  geom_hline(yintercept = 0, linetype = "dashed", linewidth = 0.4, colour = "grey55") +
  geom_line(
    data = Data_wide,
    aes(time_bin, mean, group = RKD.ID, colour = cluster_label),
    alpha = 0.15, linewidth = 0.35, show.legend = FALSE
  ) +
  geom_line(
    data = cluster_means,
    aes(time_bin, mean_value, colour = cluster_label, group = cluster_label),
    linewidth = 1.4
  ) +
  geom_point(
    data = cluster_means,
    aes(time_bin, mean_value, colour = cluster_label),
    size = 2.4
  ) +
  scale_colour_manual(values = group_cols, labels = legend_labs, drop = FALSE) +
  scale_x_continuous(
    breaks = 1:6,
    labels = c("0-1", ">1-2", ">2-3", ">3-4", ">4-5", ">5-6")
  ) +
  scale_y_continuous(
    breaks = seq(-100, 100, by = 10),
    labels = function(x) ifelse(x > 0, paste0("+", x), x)
  ) +
  coord_cartesian(ylim = c(-100, 100)) +
  labs(
    title = "Early eGFR recovery groups",
    x = "Months since diagnosis",
    y = "Change in eGFR from baseline (mL/min/1.73 m²)",
    colour = NULL
  ) +
  theme_manuscript() +
  theme(plot.title = element_text(hjust = 0.5, face = "bold", size = 15))

print(plot_c)
save_plot(plot_c, "Figure_1_egfr_recovery_groups.png", 7.2, 5.2)
```
Now we move to check whether the 5yrs-ESKD risk will differ statistically between two groups
## 5. RETRACE outcome analysis

```{r fine-gray-models}
RETRACE_cohort <- RETRACE_base %>% filter(RKD.ID %in% cluster_df$RKD.ID)

RESULT_crr <- RETRACE_cohort %>%
  arrange(RKD.ID, Interval.from.diagnosis..months.) %>%
  distinct(RKD.ID, .keep_all = TRUE) %>%
  left_join(cluster_df, by = "RKD.ID") %>%
  mutate(
    daysfrmDeath = as.numeric(dateOfDeath - dateOfDiagnosis) / 365.25,
    dateOfCensored = dateOfCensored_days / 365.25,
    daysfrmESKD = daysfrmESKD_days / 365.25,
    time_to_eskd = if_else(!is.na(daysfrmESKD), daysfrmESKD, Inf),
    time_to_death = if_else(!is.na(daysfrmDeath), daysfrmDeath, Inf),
    time_to_censor = if_else(!is.na(dateOfCensored), dateOfCensored, Inf),
    time = pmin(time_to_eskd, time_to_death, time_to_censor, na.rm = TRUE),
    status = case_when(
      ESKDStandard == "Yes" &
        time_to_eskd <= time_to_death & time_to_eskd <= time_to_censor ~ 1L,
      death == "Dead" &
        time_to_death < time_to_eskd & time_to_death <= time_to_censor ~ 2L,
      TRUE ~ 0L
    ),
    status = if_else(time > 5, 0L, status),
    time = pmin(time, 5),
    cluster_label = factor(
      cluster_label,
      levels = c("Recovered group", "Static group")
    ),
    eGFR1_10 = eGFR1 / 10,
    creatininetest1_10 = creatininetest1 / 10
  )

X <- model.matrix(~ factor(cluster), RESULT_crr)[, -1, drop = FALSE]
fit_crr <- cmprsk::crr(
  ftime = RESULT_crr$time, fstatus = RESULT_crr$status,
  cov1 = X, failcode = 1, cencode = 0, maxiter = 100
)

fg_data <- RESULT_crr %>%
  select(
    time, status, cluster, gender, ageAtDiagnosis, ancaSpec,
    inductionFinal, eGFR1_10, Berden.score.on.renal.biopsy
  ) %>%
  mutate(
    gender = factor(gender),
    ancaSpec = factor(ancaSpec, levels = c("MPO", "PR3", "Other")),
    inductionFinal = factor(inductionFinal, levels = c("Cyclo", "Rituximab", "Other")),
    Berden.score.on.renal.biopsy = factor(Berden.score.on.renal.biopsy)
  ) %>%
  filter(complete.cases(.)) %>%
  droplevels()

X_full <- model.matrix(
  ~ factor(cluster) + gender + ageAtDiagnosis + ancaSpec +
    inductionFinal + eGFR1_10 + Berden.score.on.renal.biopsy,
  fg_data
)[, -1, drop = FALSE]

fit_crr_full <- cmprsk::crr(
  ftime = fg_data$time, fstatus = fg_data$status,
  cov1 = X_full, failcode = 1, cencode = 0, maxiter = 100
)

capture.output(
  summary(fit_crr_full),
  file = file.path(output_dir, "fit_crr_full_summary.txt")
)

outcome_check <- with(
  RESULT_crr,
  table(cluster_label, factor(status, levels = 0:2))
)
print(outcome_check)
print(summary(fit_crr))
print(summary(fit_crr_full))

stopifnot(
  nrow(RESULT_crr) == 234L,
  !anyDuplicated(RESULT_crr$RKD.ID),
  all(outcome_check == matrix(c(89L, 114L, 1L, 15L, 7L, 8L), nrow = 2L))
)
```

```{r retrace-cif, fig.width=7.2, fig.height=5.35}
if (!any(RESULT_crr$status == 1L)) stop("No ESKD events available for the CIF.")

cif_fit <- cmprsk::cuminc(
  ftime = RESULT_crr$time,
  fstatus = RESULT_crr$status,
  group = RESULT_crr$cluster_label,
  cencode = 0
)

cif_df_eskd <- tidy_cuminc(cif_fit, event_code = 1L) %>%
  mutate(group = factor(group, levels = c("Recovered group", "Static group")))

fg_univariable_result <- fine_gray_stats(fit_crr)
fg_annotation <- with(
  fg_univariable_result,
  paste0(
    "Univariable Fine–Gray model\n",
    "Static group vs Recovered group: SHR ", sprintf("%.2f", SHR),
    " (95% CI ", sprintf("%.2f", lower), "–", sprintf("%.2f", upper), "); ",
    format_p(p, prefix = TRUE)
  )
)

p_cif <- ggplot(
  cif_df_eskd,
  aes(time, estimate, colour = group, fill = group)
) +
  geom_step(linewidth = 1) +
  geom_ribbon(
    aes(ymin = lower, ymax = upper),
    alpha = 0.14, colour = NA, show.legend = FALSE
  ) +
  annotate(
    "text", x = 0.20, y = 0.385, label = fg_annotation,
    hjust = 0, vjust = 1, size = 3.8, lineheight = 1.05
  ) +
  coord_cartesian(xlim = c(0, 5), ylim = c(0, 0.4), expand = FALSE) +
  scale_x_continuous(breaks = 0:5) +
  scale_y_continuous(
    breaks = seq(0, 0.4, by = 0.1),
    labels = scales::percent_format(accuracy = 1)
  ) +
  scale_colour_manual(values = group_cols, drop = FALSE) +
  scale_fill_manual(values = group_cols, drop = FALSE) +
  labs(
    title = "Cumulative incidence of ESKD",
    x = "Time in years since diagnosis",
    y = "Cumulative incidence",
    colour = NULL,
    fill = NULL
  ) +
  theme_manuscript() +
  theme(plot.title = element_text(hjust = 0.5, face = "bold", size = 15))

print(p_cif)
save_plot(p_cif, "Figure_2_ESKD_CIF.png", 7.2, 5.35)
```

After we checked the difference of CIF in two different clustering group, we now define Cohort 3 and check the eGFR slope difference between two clusters.
T1 is the earliest occurrence of the highest available eGFR between 4 and 12 months after diagnosis.Measurements from T1 through 60 months after diagnosis are retained, including measurements after ESKD; follow-up ends at transplantation, death, or last recorded contact. Eligibility requires T1 and at least one later eGFR measurement on a distinct date.


```{r post-peak-slope-data}
slope_output_dir <- file.path(output_dir, "post_peak_egfr_slope")
dir.create(slope_output_dir, recursive = TRUE, showWarnings = FALSE)

slope_ids <- cluster_df %>%
  transmute(RKD.ID, cluster_label = as.character(cluster_label)) %>%
  distinct()

patient_endpoints <- renal_raw %>%
  semi_join(slope_ids, by = "RKD.ID") %>%
  group_by(RKD.ID) %>%
  summarise(
    dateOfESKD = first_valid_date(dateOfESKD),
    dateOfDeath = first_valid_date(dateOfDeath),
    lastRecordedContact = first_valid_date(lastRecordedContact),
    .groups = "drop"
  )

transplant_dates <- if ("Date.of.transplant." %in% names(raw_data)) {
  raw_data %>%
    semi_join(slope_ids, by = "RKD.ID") %>%
    transmute(RKD.ID, dateOfTransplant = date_clean(Date.of.transplant.)) %>%
    group_by(RKD.ID) %>%
    summarise(
      dateOfTransplant = first_valid_date(sort(dateOfTransplant)),
      .groups = "drop"
    )
} else {
  slope_ids %>% transmute(RKD.ID, dateOfTransplant = as.Date(NA_character_))
}

slope_source <- renal_raw %>%
  select(
    RKD.ID, Date.Of.Visit, Interval.from.diagnosis..months.,
    egfr_calculated, Dialysis.dependent
  ) %>%
  inner_join(slope_ids, by = "RKD.ID") %>%
  left_join(patient_endpoints, by = "RKD.ID") %>%
  left_join(transplant_dates, by = "RKD.ID") %>%
  mutate(
    interval_months = num_clean(Interval.from.diagnosis..months.),
    dialysis_dependent = Dialysis.dependent == "Yes",
    eGFR = if_else(dialysis_dependent, 3, num_clean(egfr_calculated))
  ) %>%
  filter(
    !is.na(Date.Of.Visit), is.finite(interval_months), is.finite(eGFR),
    between(interval_months, 4, 60),
    is.na(dateOfTransplant) | Date.Of.Visit < dateOfTransplant,
    is.na(dateOfDeath) | Date.Of.Visit <= dateOfDeath,
    is.na(lastRecordedContact) | Date.Of.Visit <= lastRecordedContact
  )

# Collapse same-day records; any dialysis-dependent row fixes that date at 3.
same_day_audit <- slope_source %>%
  count(RKD.ID, Date.Of.Visit, name = "n_rows") %>%
  filter(n_rows > 1)

slope_visits <- slope_source %>%
  group_by(RKD.ID, Date.Of.Visit) %>%
  summarise(
    cluster_label = first(cluster_label),
    dateOfESKD = first_valid_date(dateOfESKD),
    interval_months = mean(interval_months),
    dialysis_dependent = any(dialysis_dependent),
    eGFR = if_else(dialysis_dependent, 3, mean(eGFR)),
    .groups = "drop"
  )

t1_candidates <- slope_visits %>% filter(between(interval_months, 4, 12))
t1_tie_audit <- t1_candidates %>%
  group_by(RKD.ID) %>%
  filter(eGFR == max(eGFR)) %>%
  summarise(n_tied_dates = n_distinct(Date.Of.Visit), .groups = "drop") %>%
  filter(n_tied_dates > 1)

t1_by_patient <- t1_candidates %>%
  arrange(RKD.ID, desc(eGFR), Date.Of.Visit) %>%
  group_by(RKD.ID) %>%
  slice(1) %>%
  ungroup() %>%
  transmute(
    RKD.ID, T1_date = Date.Of.Visit,
    T1_month = interval_months, T1_eGFR = eGFR
  )

t1_source_audit <- renal_raw %>%
  semi_join(slope_ids, by = "RKD.ID") %>%
  transmute(
    RKD.ID, Date.Of.Visit,
    interval_months = num_clean(Interval.from.diagnosis..months.),
    eGFR = if_else(Dialysis.dependent == "Yes", 3, num_clean(egfr_calculated))
  ) %>%
  group_by(RKD.ID) %>%
  summarise(
    any_record_4_12 = any(
      !is.na(interval_months) & between(interval_months, 4, 12)
    ),
    any_usable_eGFR_4_12 = any(
      !is.na(Date.Of.Visit) & !is.na(interval_months) &
        between(interval_months, 4, 12) & is.finite(eGFR)
    ),
    .groups = "drop"
  )

t1_disposition <- slope_ids %>%
  left_join(t1_source_audit, by = "RKD.ID") %>%
  mutate(
    across(c(any_record_4_12, any_usable_eGFR_4_12), ~ coalesce(.x, FALSE)),
    T1_status = case_when(
      RKD.ID %in% t1_by_patient$RKD.ID ~ "T1 available",
      !any_record_4_12 ~ "No record at 4-12 months",
      !any_usable_eGFR_4_12 ~ "Record present, but no usable eGFR",
      TRUE ~ "Excluded by endpoint/date rule"
    )
  ) %>%
  count(T1_status, name = "n")

slope_after_t1 <- slope_visits %>%
  inner_join(t1_by_patient, by = "RKD.ID") %>%
  filter(Date.Of.Visit >= T1_date, interval_months <= 60) %>%
  mutate(time_from_T1_years = as.numeric(Date.Of.Visit - T1_date) / 365.25)

slope_patient_eligibility <- slope_ids %>%
  left_join(t1_by_patient, by = "RKD.ID") %>%
  left_join(
    slope_after_t1 %>%
      group_by(RKD.ID) %>%
      summarise(
        n_slope_visits = n_distinct(Date.Of.Visit),
        has_post_T1_visit = any(time_from_T1_years > 0),
        maximum_followup_from_T1_years = max(time_from_T1_years),
        .groups = "drop"
      ),
    by = "RKD.ID"
  ) %>%
  mutate(
    has_T1 = !is.na(T1_date),
    n_slope_visits = coalesce(n_slope_visits, 0L),
    has_post_T1_visit = coalesce(has_post_T1_visit, FALSE),
    model_eligible = has_T1 & n_slope_visits >= 2 & has_post_T1_visit
  )

slope_flow <- bind_rows(
  slope_patient_eligibility %>%
    group_by(cluster_label) %>%
    summarise(
      RETRACE_N = n(), T1_available = sum(has_T1),
      model_N = sum(model_eligible), .groups = "drop"
    ),
  slope_patient_eligibility %>%
    summarise(
      cluster_label = "Overall", RETRACE_N = n(),
      T1_available = sum(has_T1), model_N = sum(model_eligible)
    )
)

slope_data <- slope_after_t1 %>%
  semi_join(
    slope_patient_eligibility %>% filter(model_eligible),
    by = "RKD.ID"
  ) %>%
  mutate(
    cluster_slope = factor(
      cluster_label, levels = c("Static group", "Recovered group")
    )
  ) %>%
  arrange(RKD.ID, time_from_T1_years)

dialysis3_audit <- bind_rows(
  slope_source %>% summarise(
    analysis_stage = "All eligible 4-60 month source rows",
    dialysis_rows = sum(dialysis_dependent),
    dialysis_patients = n_distinct(RKD.ID[dialysis_dependent])
  ),
  slope_data %>% summarise(
    analysis_stage = "Primary model data from T1 onward",
    dialysis_rows = sum(dialysis_dependent),
    dialysis_patients = n_distinct(RKD.ID[dialysis_dependent])
  )
)

stopifnot(
  nrow(slope_ids) == 234L,
  slope_flow$T1_available[slope_flow$cluster_label == "Overall"] == 166L,
  slope_flow$model_N[slope_flow$cluster_label == "Overall"] == 148L,
  sum(slope_patient_eligibility$model_eligible &
        slope_patient_eligibility$cluster_label == "Recovered group") == 58L,
  sum(slope_patient_eligibility$model_eligible &
        slope_patient_eligibility$cluster_label == "Static group") == 90L,
  nrow(slope_data) == 875L
)

print(slope_flow)
print(t1_disposition)
print(dialysis3_audit)
```

```{r post-peak-slope-model-and-figures, fig.width=10.2, fig.height=9.8}
slope_control <- nlme::lmeControl(
  opt = "optim", maxIter = 200, msMaxIter = 200,
  niterEM = 50, returnObject = TRUE
)
slope_model <- nlme::lme(
  eGFR ~ time_from_T1_years * cluster_slope,
  random = ~ time_from_T1_years | RKD.ID,
  data = slope_data, method = "REML", na.action = na.exclude,
  control = slope_control
)

# Sensitivity checks for the random-slope and linear-time specifications.
slope_random_intercept_model <- nlme::lme(
  eGFR ~ time_from_T1_years * cluster_slope,
  random = ~ 1 | RKD.ID,
  data = slope_data, method = "REML", na.action = na.exclude,
  control = slope_control
)
slope_random_effect_comparison <- anova(
  slope_random_intercept_model, slope_model
)
slope_linear_ml_model <- update(slope_model, method = "ML")
slope_quadratic_ml_model <- nlme::lme(
  eGFR ~ (time_from_T1_years + I(time_from_T1_years^2)) * cluster_slope,
  random = ~ time_from_T1_years | RKD.ID,
  data = slope_data, method = "ML", na.action = na.exclude,
  control = slope_control
)
slope_linearity_comparison <- anova(
  slope_linear_ml_model, slope_quadratic_ml_model
)

slope_beta <- nlme::fixef(slope_model)
slope_vcov <- vcov(slope_model)
slope_time_term <- "time_from_T1_years"
slope_interaction_term <- grep(
  "time_from_T1_years:cluster_slope|cluster_slope.*:time_from_T1_years",
  names(slope_beta), value = TRUE
)
stopifnot(length(slope_interaction_term) == 1L)
slope_df <- min(summary(slope_model)$tTable[
  c(slope_time_term, slope_interaction_term), "DF"
])

slope_contrast <- function(weights, label) {
  L <- setNames(rep(0, length(slope_beta)), names(slope_beta))
  L[names(weights)] <- weights
  estimate <- sum(L * slope_beta)
  SE <- sqrt(as.numeric(t(L) %*% slope_vcov %*% L))
  critical <- qt(0.975, slope_df)
  tibble(
    estimand = label, estimate, SE, df = slope_df,
    lower_95_CI = estimate - critical * SE,
    upper_95_CI = estimate + critical * SE,
    p_value = 2 * pt(abs(estimate / SE), slope_df, lower.tail = FALSE)
  )
}

slope_results <- bind_rows(
  slope_contrast(
    setNames(1, slope_time_term),
    "Static group annual post-peak eGFR slope"
  ),
  slope_contrast(
    setNames(c(1, 1), c(slope_time_term, slope_interaction_term)),
    "Recovered group annual post-peak eGFR slope"
  ),
  slope_contrast(
    setNames(1, slope_interaction_term),
    "Recovered minus Static slope difference"
  )
) %>%
  mutate(
    unit = "mL/min/1.73 m^2/year",
    estimate_95_CI = sprintf(
      "%.2f (%.2f to %.2f)", estimate, lower_95_CI, upper_95_CI
    ),
    p_value_formatted = vapply(p_value, format_p, character(1))
  )

fixed_effect_table <- as.data.frame(summary(slope_model)$tTable) %>%
  tibble::rownames_to_column("term")

prediction_grid <- expand_grid(
  time_from_T1_years = seq(0, max(slope_data$time_from_T1_years), length.out = 120),
  cluster_slope = factor(
    c("Static group", "Recovered group"),
    levels = c("Static group", "Recovered group")
  )
)
slope_design <- model.matrix(
  ~ time_from_T1_years * cluster_slope, prediction_grid
)
prediction_grid <- prediction_grid %>%
  mutate(
    estimate = as.numeric(slope_design %*% slope_beta),
    SE = sqrt(rowSums((slope_design %*% slope_vcov) * slope_design)),
    lower = estimate - 1.96 * SE,
    upper = estimate + 1.96 * SE
  )

slope_legend <- slope_data %>%
  distinct(RKD.ID, cluster_slope) %>%
  count(cluster_slope, name = "n") %>%
  transmute(
    cluster_slope,
    label = paste0(cluster_slope, " (n = ", n, ")")
  ) %>%
  { setNames(.$label, .$cluster_slope) }

p_slope_trajectory <- ggplot(
  slope_data,
  aes(time_from_T1_years, eGFR, group = RKD.ID, colour = cluster_slope)
) +
  geom_line(alpha = 0.075, linewidth = 0.28) +
  geom_point(alpha = 0.12, size = 0.62) +
  geom_ribbon(
    data = prediction_grid,
    aes(
      x = time_from_T1_years, ymin = lower, ymax = upper,
      fill = cluster_slope, group = cluster_slope
    ),
    inherit.aes = FALSE, alpha = 0.16, colour = NA
  ) +
  geom_line(
    data = prediction_grid,
    aes(
      time_from_T1_years, estimate,
      colour = cluster_slope, group = cluster_slope
    ),
    inherit.aes = FALSE, linewidth = 1.18
  ) +
  scale_colour_manual(
    values = group_cols,
    breaks = c("Recovered group", "Static group"),
    labels = slope_legend[c("Recovered group", "Static group")]
  ) +
  scale_fill_manual(values = group_cols) +
  scale_x_continuous(
    breaks = 0:5, limits = c(0, max(slope_data$time_from_T1_years)),
    expand = expansion(mult = c(0.01, 0.025))
  ) +
  scale_y_continuous(breaks = c(0, 20, 40, 60, 80, 90)) +
  coord_cartesian(ylim = c(0, 90), expand = FALSE) +
  labs(
    title = "Post-peak eGFR trajectories by recovery group",
    x = "Years since T1",
    y = expression(eGFR~"(mL/min/1.73"~m^2*")"),
    colour = NULL, fill = NULL
  ) +
  theme_classic(base_size = 11.5) +
  theme(
    plot.title = element_text(face = "bold", size = 13.5),
    axis.title = element_text(face = "bold"),
    axis.text = element_text(colour = "black"),
    legend.position = "bottom",
    legend.justification = "left",
    legend.key.width = grid::unit(1.6, "lines")
  ) +
  guides(
    fill = "none",
    colour = guide_legend(override.aes = list(alpha = 1, linewidth = 1.5))
  )

forest_data <- slope_results %>%
  mutate(
    row = c(3, 2, 1),
    label = c(
      "Static group slope", "Recovered group slope",
      "Recovered - Static difference"
    ),
    series = c("Static group", "Recovered group", "Difference"),
    result_label = c(
      sprintf(
        "%.2f (%.2f to %.2f); P = %.3f",
        estimate[1], lower_95_CI[1], upper_95_CI[1], p_value[1]
      ),
      sprintf(
        "%.2f (%.2f to %.2f); P = %.3f",
        estimate[2], lower_95_CI[2], upper_95_CI[2], p_value[2]
      ),
      sprintf(
        "%.2f (%.2f to %.2f); interaction P = %.3f",
        estimate[3], lower_95_CI[3], upper_95_CI[3], p_value[3]
      )
    )
  )

p_slope_forest <- ggplot(forest_data, aes(y = row)) +
  geom_vline(xintercept = 0, linetype = "dashed", colour = "grey45") +
  geom_segment(
    aes(
      x = lower_95_CI, xend = upper_95_CI,
      yend = row, colour = series
    ),
    linewidth = 1.05, lineend = "round"
  ) +
  geom_point(aes(x = estimate, colour = series, shape = series), size = 3.4) +
  geom_text(
    aes(x = 5.35, label = result_label),
    hjust = 0, size = 3.35, colour = "black"
  ) +
  annotate(
    "text", x = 5.35, y = 3.72,
    label = "Estimate (95% CI); P value",
    hjust = 0, fontface = "bold", size = 3.45
  ) +
  scale_colour_manual(
    values = c(group_cols, "Difference" = "#222222"), guide = "none"
  ) +
  scale_shape_manual(
    values = c("Static group" = 16, "Recovered group" = 16, "Difference" = 18),
    guide = "none"
  ) +
  scale_y_continuous(
    breaks = forest_data$row, labels = forest_data$label,
    limits = c(0.55, 3.85), expand = expansion(mult = c(0, 0))
  ) +
  scale_x_continuous(breaks = seq(-4, 4, 2), expand = expansion(mult = c(0, 0))) +
  coord_cartesian(xlim = c(-4.3, 5.15), clip = "off") +
  labs(
    title = "Annual post-peak eGFR slopes",
    x = expression(
      "Annual eGFR change or difference (mL/min/1.73"~m^2*"/year)"
    ),
    y = NULL
  ) +
  theme_classic(base_size = 11.5) +
  theme(
    plot.title = element_text(face = "bold", size = 13.5),
    axis.title.x = element_text(face = "bold"),
    axis.text = element_text(colour = "black"),
    axis.line.y = element_blank(), axis.ticks.y = element_blank(),
    plot.margin = margin(8, 245, 10, 12)
  )

p_slope_combined <- (
  p_slope_trajectory / p_slope_forest +
    patchwork::plot_layout(heights = c(1.18, 1)) +
    patchwork::plot_annotation(tag_levels = "A")
) & theme(plot.tag = element_text(face = "bold", size = 15))

print(slope_results %>% select(estimand, estimate_95_CI, p_value_formatted))
print(slope_random_effect_comparison)
print(slope_linearity_comparison)
print(nlme::VarCorr(slope_model))
print(p_slope_combined)

save_table(slope_flow, "post_peak_egfr_slope/RETRACE_post_peak_slope_cohort_flow.csv")
save_table(t1_disposition, "post_peak_egfr_slope/RETRACE_post_peak_T1_disposition.csv")
save_table(t1_by_patient, "post_peak_egfr_slope/RETRACE_post_peak_T1_by_patient.csv")
save_table(dialysis3_audit, "post_peak_egfr_slope/RETRACE_post_peak_dialysis3_audit.csv")
save_table(
  slope_patient_eligibility,
  "post_peak_egfr_slope/RETRACE_post_peak_slope_patient_eligibility.csv"
)
save_table(same_day_audit, "post_peak_egfr_slope/RETRACE_post_peak_same_day_audit.csv")
save_table(t1_tie_audit, "post_peak_egfr_slope/RETRACE_post_peak_T1_tie_audit.csv")
save_table(slope_results, "post_peak_egfr_slope/RETRACE_post_peak_slope_results.csv")
save_table(
  fixed_effect_table,
  "post_peak_egfr_slope/RETRACE_post_peak_slope_fixed_effects.csv"
)
save_table(
  prediction_grid,
  "post_peak_egfr_slope/RETRACE_post_peak_slope_fitted_curves.csv"
)
save_table(
  as.data.frame(slope_random_effect_comparison),
  "post_peak_egfr_slope/RETRACE_post_peak_random_effect_comparison.csv"
)
save_table(
  as.data.frame(slope_linearity_comparison),
  "post_peak_egfr_slope/RETRACE_post_peak_linearity_comparison.csv"
)
saveRDS(slope_model, file.path(slope_output_dir, "RETRACE_post_peak_slope_model.rds"))
save_plot(
  p_slope_trajectory,
  "post_peak_egfr_slope/Figure_post_peak_eGFR_trajectories.png", 10.2, 5.35
)
save_plot(
  p_slope_forest,
  "post_peak_egfr_slope/Figure_post_peak_eGFR_slope_forest.png", 10.2, 4.5
)
save_plot(
  p_slope_combined,
  "post_peak_egfr_slope/Figure_post_peak_eGFR_combined.png", 10.2, 9.8
)
save_plot(
  p_slope_combined,
  "post_peak_egfr_slope/Figure_post_peak_eGFR_combined.pdf", 10.2, 9.8
)
```

### Cohort 3 characteristics

```{r cohort3-characteristics-table}
cohort3_ids <- slope_patient_eligibility %>%
  filter(model_eligible) %>%
  select(
    RKD.ID, cluster_label, T1_date, T1_month, T1_eGFR,
    n_slope_visits, maximum_followup_from_T1_years
  )

# Reuse the Cohort 2 baseline rule and its dialysis substitutions.
cohort3_baseline <- bind_rows(baseline_post, baseline_pre) %>%
  distinct(RKD.ID, .keep_all = TRUE) %>%
  transmute(
    RKD.ID,
    baseline_visit_date = Date.Of.Visit,
    baseline_month = Interval.from.diagnosis..months.,
    baseline_creatinine = creatininetest,
    baseline_eGFR = egfr_calculated,
    baseline_dialysis_dependent = Dialysis.dependent == "Yes"
  )

any_flagged <- function(x, yes = "Checked") {
  x <- trimws(as.character(x))
  if (any(x %in% yes, na.rm = TRUE)) TRUE else
    if (all(is.na(x) | x == "")) NA else FALSE
}

cohort3_features <- raw_data %>%
  semi_join(cohort3_ids, by = "RKD.ID") %>%
  group_by(RKD.ID) %>%
  summarise(
    age_at_diagnosis = first_finite(ageAtDiagnosis),
    sex = first_nonblank(gender),
    anca_specificity = {
      x <- first_nonblank(ancaSpec)
      if (is.na(x)) NA_character_ else if (x %in% c("MPO", "PR3")) x else "Other"
    },
    diagnosis = case_when(
      any(num_clean(DefiniteMPA) == 1, na.rm = TRUE) ~ "MPA",
      any(num_clean(DefiniteGPA) == 1, na.rm = TRUE) ~ "GPA",
      TRUE ~ NA_character_
    ),
    berden_class = {
      x <- trimws(as.character(Berden.score.on.renal.biopsy))
      x <- x[x %in% c("Crescentic", "Focal", "Mixed", "Sclerotic")]
      if (length(x)) x[1] else "Berden class unavailable"
    },
    rituximab_used = any_flagged(inductiontreatmentType_Rituximab),
    cyclophosphamide_used = any(
      trimws(as.character(inductiontreatmentType_Cyclophosphamide)) == "Checked" |
        trimws(as.character(
          inductiontreatmentType_.Daily.Oral.Cyclophosphamide
        )) == "Checked",
      na.rm = TRUE
    ),
    oral_corticosteroids = any_flagged(inductiontreatmentType_Oral.corticosteroids),
    pulsed_iv_corticosteroids = any_flagged(
      inductiontreatmentType_.Pulsed.IV.corticosteroids
    ),
    plasma_exchange = any_flagged(inductiontreatmentType_Plasma.exchange),
    across(all_of(organ_cols), any_flagged),
    .groups = "drop"
  ) %>%
  mutate(
    induction_treatment = case_when(
      cyclophosphamide_used & !rituximab_used ~ "Cyclophosphamide",
      rituximab_used & !cyclophosphamide_used ~ "Rituximab",
      TRUE ~ "Other"
    )
  )

cohort3 <- cohort3_ids %>%
  left_join(cohort3_features, by = "RKD.ID") %>%
  left_join(cohort3_baseline, by = "RKD.ID") %>%
  mutate(
    cluster_label = factor(
      cluster_label, levels = c("Recovered group", "Static group")
    ),
    sex = factor(sex, levels = c("Female", "Male")),
    anca_specificity = factor(
      anca_specificity, levels = c("MPO", "PR3", "Other")
    ),
    diagnosis = factor(diagnosis, levels = c("MPA", "GPA")),
    berden_class = factor(
      berden_class,
      levels = c(
        "Crescentic", "Focal", "Mixed", "Sclerotic",
        "Berden class unavailable"
      )
    ),
    induction_treatment = factor(
      induction_treatment,
      levels = c("Cyclophosphamide", "Rituximab", "Other")
    )
  )

cohort3_groups <- c("Overall", "Recovered group", "Static group")
cohort3_n <- c(
  Overall = nrow(cohort3),
  `Recovered group` = sum(cohort3$cluster_label == "Recovered group"),
  `Static group` = sum(cohort3$cluster_label == "Static group")
)
cohort3_group_columns <- setNames(
  sprintf("%s (N = %d)", cohort3_groups, cohort3_n[cohort3_groups]),
  cohort3_groups
)

c3_subset <- function(group) {
  if (group == "Overall") cohort3 else filter(cohort3, cluster_label == group)
}
c3_median_iqr <- function(data, variable, digits = 1) {
  x <- data[[variable]]
  x <- x[is.finite(x)]
  if (!length(x)) return("")
  q <- quantile(x, c(0.25, 0.5, 0.75), names = FALSE)
  sprintf(
    paste0("%.", digits, "f (%.", digits, "f-%.", digits, "f)"),
    q[2], q[1], q[3]
  )
}
c3_n_pct <- function(data, variable, level = TRUE, denominator = NULL) {
  x <- data[[variable]]
  if (is.null(denominator)) denominator <- sum(!is.na(x))
  n <- sum(x == level, na.rm = TRUE)
  if (!denominator) return("")
  sprintf("%d/%d (%.1f%%)", n, denominator, 100 * n / denominator)
}
c3_p <- function(p) if (is.na(p)) "" else format_p(p)
c3_values <- function(formatter) {
  formatter <- rlang::as_function(formatter)
  setNames(
    vapply(cohort3_groups, function(g) formatter(c3_subset(g)), character(1)),
    cohort3_group_columns
  )
}
c3_row <- function(section, variable, values, p = "") {
  bind_cols(
    tibble(Section = section, Variable = variable),
    tibble::as_tibble_row(values),
    tibble(`P value` = p)
  )
}
c3_continuous <- function(section, label, variable, digits = 1) {
  c3_row(
    section, label,
    c3_values(~ c3_median_iqr(.x, variable, digits)),
    c3_p(wilcox_p(cohort3[[variable]], cohort3$cluster_label))
  )
}
c3_binary <- function(section, label, variable, level = TRUE) {
  c3_row(
    section, label,
    c3_values(~ c3_n_pct(.x, variable, level)),
    c3_p(fisher_p(cohort3[[variable]], cohort3$cluster_label))
  )
}
c3_factor_header <- function(section, label, variable, data = cohort3) {
  c3_row(
    section, label, setNames(rep("", 3), cohort3_group_columns),
    c3_p(fisher_p(data[[variable]], data$cluster_label))
  )
}
c3_factor_level <- function(section, label, variable, level) {
  c3_row(
    section, paste0("  ", label),
    c3_values(~ c3_n_pct(.x, variable, level))
  )
}
c3_berden <- function(data, level) {
  denominator <- if (level == "Berden class unavailable") {
    nrow(data)
  } else {
    sum(!is.na(data$berden_class) &
          data$berden_class != "Berden class unavailable")
  }
  c3_n_pct(data, "berden_class", level, denominator)
}

c3_berden_available <- cohort3 %>%
  filter(berden_class != "Berden class unavailable") %>%
  droplevels()
c3_has_berden <- cohort3$berden_class != "Berden class unavailable"

cohort3_table <- bind_rows(
  c3_continuous("Demographic characteristics", "Age at diagnosis, years", "age_at_diagnosis"),
  c3_binary("Demographic characteristics", "Female sex", "sex", "Female"),
  c3_factor_header("Disease characteristics", "Clinical diagnosis", "diagnosis"),
  c3_factor_level("Disease characteristics", "MPA", "diagnosis", "MPA"),
  c3_factor_level("Disease characteristics", "GPA", "diagnosis", "GPA"),
  c3_factor_header("Disease characteristics", "ANCA specificity", "anca_specificity"),
  c3_factor_level("Disease characteristics", "MPO-ANCA", "anca_specificity", "MPO"),
  c3_factor_level("Disease characteristics", "PR3-ANCA", "anca_specificity", "PR3"),
  c3_factor_level("Disease characteristics", "Other ANCA result*", "anca_specificity", "Other"),
  c3_binary("Organ/system involvement (ever recorded)", "Constitutional", "affectedOrgan_General"),
  c3_binary("Organ/system involvement (ever recorded)", "Cutaneous", "affectedOrgan_Cutaneous"),
  c3_binary("Organ/system involvement (ever recorded)", "Mucous membrane/ocular", "affectedOrgan_Mucous_membranes_eyes"),
  c3_binary("Organ/system involvement (ever recorded)", "Ear, nose, and throat", "affectedOrgan_ENT"),
  c3_binary("Organ/system involvement (ever recorded)", "Pulmonary", "affectedOrgan_Chest"),
  c3_binary("Organ/system involvement (ever recorded)", "Cardiovascular", "affectedOrgan_Cardiovascular"),
  c3_binary("Organ/system involvement (ever recorded)", "Gastrointestinal", "affectedOrgan_Abdominal"),
  c3_binary("Organ/system involvement (ever recorded)", "Neurological", "affectedOrgan_Nervous_system"),
  c3_binary("Organ/system involvement (ever recorded)", "Renal", "affectedOrgan_Renal"),
  c3_binary("Organ/system involvement (ever recorded)", "Other recorded system involvement**", "affectedOrgan_Other"),
  c3_continuous("Kidney characteristics", "Baseline serum creatinine, µmol/L", "baseline_creatinine", 0),
  c3_continuous("Kidney characteristics", "Baseline eGFR, mL/min/1.73 m²", "baseline_eGFR", 0),
  c3_binary("Kidney characteristics", "Dialysis-dependent at baseline", "baseline_dialysis_dependent"),
  c3_factor_header(
    "Kidney histology", "Berden histological class", "berden_class",
    c3_berden_available
  ),
  c3_row("Kidney histology", "  Crescentic", c3_values(~ c3_berden(.x, "Crescentic"))),
  c3_row("Kidney histology", "  Focal", c3_values(~ c3_berden(.x, "Focal"))),
  c3_row("Kidney histology", "  Mixed", c3_values(~ c3_berden(.x, "Mixed"))),
  c3_row("Kidney histology", "  Sclerotic", c3_values(~ c3_berden(.x, "Sclerotic"))),
  c3_row(
    "Kidney histology", "Berden class unavailable",
    c3_values(~ c3_berden(.x, "Berden class unavailable")),
    c3_p(fisher_p(c3_has_berden, cohort3$cluster_label))
  ),
  c3_factor_header("Induction treatment", "Recorded induction regimen", "induction_treatment"),
  c3_factor_level("Induction treatment", "Cyclophosphamide only", "induction_treatment", "Cyclophosphamide"),
  c3_factor_level("Induction treatment", "Rituximab only", "induction_treatment", "Rituximab"),
  c3_factor_level("Induction treatment", "Combined CYC + RTX or other induction***", "induction_treatment", "Other"),
  c3_binary("Induction treatment", "Oral corticosteroids", "oral_corticosteroids"),
  c3_binary("Induction treatment", "Pulsed intravenous corticosteroids", "pulsed_iv_corticosteroids"),
  c3_binary("Induction treatment", "Plasma exchange", "plasma_exchange"),
  c3_continuous("Post-peak analysis", "T1, months after diagnosis", "T1_month"),
  c3_continuous("Post-peak analysis", "eGFR at T1, mL/min/1.73 m²", "T1_eGFR", 0),
  c3_continuous("Post-peak analysis", "Included eGFR measurement dates from T1, n", "n_slope_visits", 0),
  c3_continuous("Post-peak analysis", "Span from T1 to last included eGFR measurement, years", "maximum_followup_from_T1_years")
)

cohort3_notes <- c(
  "Characteristics of Cohort 3 (the post-peak eGFR slope analysis subset of Cohort 2), overall and by early eGFR recovery group.",
  "Values are median (interquartile range) or n/N (%), where N is the number with available data unless otherwise stated.",
  "For dialysis-dependent visits, eGFR was assigned 3 mL/min/1.73 m² before baseline and T1 selection; serum creatinine was assigned 500 µmol/L before baseline selection.",
  "The four Berden-class percentages use patients with an available class as the denominator (overall N=110; Recovered N=41; Static N=69); the unavailable row uses the full group denominator.",
  "* Other ANCA result comprises two ELISA-negative and two dual MPO/PR3 results.",
  "** Other recorded system involvement comprises peripheral vascular disease with gangrenous toes/amputation, a constitutional entry, and a mediastinal mass.",
  "*** This category comprises 13 patients who received both cyclophosphamide and rituximab and four who received another induction agent (three azathioprine and one mycophenolate mofetil).",
  "T1 is the earliest occurrence of the highest available eGFR between 4 and 12 months after diagnosis, after endpoint/date filtering and same-day record collapse.",
  "Slope-model eligibility required T1 and at least one later eGFR measurement on a distinct date; measurements from T1 through 60 months after diagnosis were included.",
  "P values compare the Recovered and Static groups using Wilcoxon rank-sum or Fisher's exact tests; they are descriptive, unadjusted, and were not corrected for multiple comparisons."
)

save_table(cohort3, "post_peak_egfr_slope/cohort3_post_peak_patient_level.csv")
save_table(
  cohort3_table,
  "post_peak_egfr_slope/cohort3_post_peak_characteristics_table.csv"
)
save_table(
  cohort3_table %>% filter(Section != "Post-peak analysis"),
  "post_peak_egfr_slope/cohort3_post_peak_baseline_characteristics.csv"
)
save_table(
  cohort3_table %>% filter(Section == "Post-peak analysis"),
  "post_peak_egfr_slope/cohort3_post_peak_analysis_characteristics.csv"
)
writeLines(
  cohort3_notes,
  file.path(slope_output_dir, "cohort3_post_peak_table_notes.txt"),
  useBytes = TRUE
)

stopifnot(
  nrow(cohort3) == 148L,
  sum(cohort3$cluster_label == "Recovered group") == 58L,
  sum(cohort3$cluster_label == "Static group") == 90L,
  nrow(cohort3_table) == 39L,
  !any(is.na(cohort3$baseline_eGFR))
)
print(cohort3_table, n = Inf, width = Inf)
```
```















## 6. Supplementary full and RETRACE-excluded CIFs

```{r supplementary-full-and-excluded-cif, fig.width=12, fig.height=5.5}
# This analysis is descriptive: the excluded cohort is a subset of the full
# cohort, so the panels must not be compared with a between-group test.
cif_followup_years <- 5
prediagnosis_eskd_policy <- "time_zero"
stopifnot(prediagnosis_eskd_policy %in% c("time_zero", "exclude", "stop"))

cif_output_dir <- file.path(output_dir, "cif_full_and_excluded")
dir.create(cif_output_dir, recursive = TRUE, showWarnings = FALSE)

full_cif_fields <- c(
  "dateOfDiagnosis", "dateOfESKD", "dateOfDeath",
  "lastRecordedContact", "ESKDStandard", "death"
)
assert_columns(Full_cohort, full_cif_fields)

# Confirm patient-level dates/statuses do not conflict across visit rows.
full_cif_field_audit <- Full_cohort %>%
  filter(!is.na(RKD.ID)) %>%
  group_by(RKD.ID) %>%
  summarise(
    across(
      all_of(full_cif_fields),
      ~ n_distinct(.x[!is.na(.x) & trimws(as.character(.x)) != ""]),
      .names = "n_values_{.col}"
    ),
    .groups = "drop"
  )

full_cif_field_conflicts <- full_cif_field_audit %>%
  filter(if_any(starts_with("n_values_"), ~ .x > 1))
write.csv(
  full_cif_field_conflicts,
  file.path(cif_output_dir, "CIF_patient_field_conflicts.csv"),
  row.names = FALSE
)
if (nrow(full_cif_field_conflicts)) {
  stop("Conflicting patient-level CIF fields; review CIF_patient_field_conflicts.csv.")
}

full_cif_patient <- Full_cohort %>%
  filter(!is.na(RKD.ID)) %>%
  arrange(RKD.ID, Interval.from.diagnosis..months., Date.Of.Visit) %>%
  group_by(RKD.ID) %>%
  summarise(across(all_of(full_cif_fields), first_nonblank), .groups = "drop")

retrace_cif_ids <- RESULT_crr$RKD.ID
stopifnot(
  !anyDuplicated(retrace_cif_ids),
  all(retrace_cif_ids %in% full_cif_patient$RKD.ID)
)

full_cif_patient <- full_cif_patient %>%
  mutate(
    trajectory_eligibility = factor(
      if_else(RKD.ID %in% retrace_cif_ids, "RETRACE included", "RETRACE excluded"),
      levels = c("RETRACE included", "RETRACE excluded")
    ),
    eskd_flag = coalesce(toupper(trimws(ESKDStandard)) == "YES", FALSE),
    death_flag = coalesce(toupper(trimws(death)) == "DEAD", FALSE),
    days_to_eskd_raw = as.numeric(dateOfESKD - dateOfDiagnosis),
    days_to_death_raw = as.numeric(dateOfDeath - dateOfDiagnosis),
    days_to_censor_raw = as.numeric(lastRecordedContact - dateOfDiagnosis),
    prediagnosis_eskd = eskd_flag & !is.na(days_to_eskd_raw) & days_to_eskd_raw < 0,
    eskd_after_last_contact = eskd_flag &
      !is.na(dateOfESKD) & !is.na(lastRecordedContact) &
      dateOfESKD > lastRecordedContact
  )

prediagnosis_eskd_records <- full_cif_patient %>%
  filter(prediagnosis_eskd) %>%
  select(
    RKD.ID, trajectory_eligibility,
    dateOfDiagnosis, dateOfESKD, days_to_eskd_raw
  )
write.csv(
  prediagnosis_eskd_records,
  file.path(cif_output_dir, "CIF_prediagnosis_ESKD_audit.csv"),
  row.names = FALSE
)
if (prediagnosis_eskd_policy == "stop" && nrow(prediagnosis_eskd_records)) {
  stop("ESKD dates before diagnosis require adjudication.")
}

full_cif_patient <- full_cif_patient %>%
  mutate(
    days_to_eskd_used = case_when(
      prediagnosis_eskd & prediagnosis_eskd_policy == "time_zero" ~ 0,
      prediagnosis_eskd ~ NA_real_,
      TRUE ~ days_to_eskd_raw
    ),
    time_to_eskd_days = if_else(
      eskd_flag & !is.na(days_to_eskd_used) & days_to_eskd_used >= 0,
      days_to_eskd_used, Inf
    ),
    time_to_death_days = if_else(
      death_flag & !is.na(days_to_death_raw) & days_to_death_raw >= 0,
      days_to_death_raw, Inf
    ),
    time_to_censor_days = if_else(
      !is.na(days_to_censor_raw) & days_to_censor_raw >= 0,
      days_to_censor_raw, Inf
    ),
    first_time_days = pmin(
      time_to_eskd_days, time_to_death_days, time_to_censor_days
    ),
    exclusion_reason = case_when(
      is.na(dateOfDiagnosis) ~ "Missing diagnosis date",
      eskd_flag & is.na(dateOfESKD) ~ "ESKD status without ESKD date",
      death_flag & is.na(dateOfDeath) ~ "Death status without death date",
      !is.na(days_to_death_raw) & days_to_death_raw < 0 ~ "Death date before diagnosis",
      !is.na(days_to_censor_raw) & days_to_censor_raw < 0 ~ "Last contact before diagnosis",
      prediagnosis_eskd & prediagnosis_eskd_policy == "exclude" ~ "ESKD before diagnosis",
      !is.finite(first_time_days) ~ "No valid outcome follow-up",
      TRUE ~ NA_character_
    ),
    status_raw = case_when(
      !is.finite(first_time_days) ~ NA_integer_,
      time_to_eskd_days <= time_to_death_days &
        time_to_eskd_days <= time_to_censor_days ~ 1L,
      time_to_death_days < time_to_eskd_days &
        time_to_death_days <= time_to_censor_days ~ 2L,
      TRUE ~ 0L
    ),
    status = if_else(
      first_time_days > cif_followup_years * 365.25, 0L, status_raw
    ),
    time = pmin(first_time_days / 365.25, cif_followup_years)
  )

full_cif_quality_audit <- full_cif_patient %>%
  filter(
    is.na(dateOfDiagnosis) | prediagnosis_eskd | eskd_after_last_contact |
      (!is.na(days_to_death_raw) & days_to_death_raw < 0) |
      (!is.na(days_to_censor_raw) & days_to_censor_raw < 0) |
      !is.na(exclusion_reason)
  ) %>%
  select(
    RKD.ID, trajectory_eligibility,
    dateOfDiagnosis, dateOfESKD, dateOfDeath, lastRecordedContact,
    days_to_eskd_raw, days_to_death_raw, days_to_censor_raw,
    prediagnosis_eskd, eskd_after_last_contact, exclusion_reason
  )

full_cif_disposition <- full_cif_patient %>%
  mutate(CIF_disposition = coalesce(exclusion_reason, "Included in CIF")) %>%
  count(trajectory_eligibility, CIF_disposition, name = "n")

write.csv(
  full_cif_quality_audit,
  file.path(cif_output_dir, "CIF_patient_level_data_quality_audit.csv"),
  row.names = FALSE
)
write.csv(
  full_cif_disposition,
  file.path(cif_output_dir, "CIF_analysis_disposition.csv"),
  row.names = FALSE
)

full_cif_analysis <- full_cif_patient %>%
  filter(is.na(exclusion_reason), is.finite(time), !is.na(status))
excluded_cif_analysis <- full_cif_analysis %>%
  filter(trajectory_eligibility == "RETRACE excluded")

fit_single_cif <- function(data) {
  if (!any(data$status == 1L)) stop("A plotted cohort has no ESKD events.")
  cmprsk::cuminc(data$time, data$status, cencode = 0)
}

full_cif_curves <- bind_rows(
  tidy_cuminc(fit_single_cif(full_cif_analysis), 1L, "Full cohort"),
  tidy_cuminc(fit_single_cif(excluded_cif_analysis), 1L, "RETRACE excluded")
)

full_cif_summary <- extract_cif_times(full_cif_curves) %>%
  mutate(across(c(estimate, lower, upper), ~ 100 * .x, .names = "{.col}_percent"))

full_cif_outcome_counts <- bind_rows(
  full_cif_analysis %>%
    summarise(
      population = "Full cohort", N = n(),
      ESKD_events = sum(status == 1L),
      competing_deaths = sum(status == 2L),
      censored = sum(status == 0L)
    ),
  full_cif_analysis %>%
    group_by(trajectory_eligibility) %>%
    summarise(
      population = as.character(first(trajectory_eligibility)), N = n(),
      ESKD_events = sum(status == 1L),
      competing_deaths = sum(status == 2L),
      censored = sum(status == 0L),
      .groups = "drop"
    ) %>%
    select(-trajectory_eligibility)
)

write.csv(
  full_cif_curves %>% select(population, time, estimate, lower, upper),
  file.path(cif_output_dir, "CIF_curve_coordinates.csv"),
  row.names = FALSE
)
write.csv(
  full_cif_summary,
  file.path(cif_output_dir, "CIF_summary_1_3_5_years.csv"),
  row.names = FALSE
)
write.csv(
  full_cif_outcome_counts,
  file.path(cif_output_dir, "CIF_outcome_counts.csv"),
  row.names = FALSE
)

full_source_n <- nrow(full_cif_patient)
full_analysis_n <- nrow(full_cif_analysis)
excluded_source_n <- sum(full_cif_patient$trajectory_eligibility == "RETRACE excluded")
excluded_analysis_n <- nrow(excluded_cif_analysis)

cif_y_upper <- min(
  1,
  max(0.25, ceiling(max(full_cif_curves$upper, na.rm = TRUE) / 0.05) * 0.05)
)

make_cif_panel <- function(population_value, colour, title) {
  ggplot(
    full_cif_curves %>% filter(.data$population == .env$population_value),
    aes(time, estimate)
  ) +
    geom_ribbon(
      aes(ymin = lower, ymax = upper),
      fill = colour, alpha = 0.14, colour = NA
    ) +
    geom_step(linewidth = 1.05, colour = colour) +
    coord_cartesian(
      xlim = c(0, cif_followup_years),
      ylim = c(0, cif_y_upper),
      expand = FALSE
    ) +
    scale_x_continuous(breaks = 0:cif_followup_years) +
    scale_y_continuous(
      breaks = seq(0, cif_y_upper, by = 0.05),
      labels = scales::percent_format(accuracy = 1)
    ) +
    labs(
      title = title,
      x = "Time in years since diagnosis",
      y = "Cumulative incidence of ESKD"
    ) +
    theme_manuscript(base_size = 12, legend = "none") +
    theme(plot.title = element_text(face = "bold", size = 13, hjust = 0))
}

p_full_cohort_cif <- make_cif_panel(
  "Full cohort", "grey20",
  sprintf("A. Full cohort (n = %d/%d analyzable)", full_analysis_n, full_source_n)
)
p_excluded_cohort_cif <- make_cif_panel(
  "RETRACE excluded", "#D55E00",
  sprintf(
    "B. RETRACE-excluded cohort (n = %d/%d analyzable)",
    excluded_analysis_n, excluded_source_n
  )
)

combined_full_excluded_cif <-
  (p_full_cohort_cif | p_excluded_cohort_cif) +
  plot_annotation(
    title = "Cumulative incidence of ESKD",
    subtitle = "Death was treated as a competing event",
    caption = paste0(
      nrow(prediagnosis_eskd_records),
      " pre-diagnosis ESKD records are displayed as time-zero events; ",
      sum(!is.na(full_cif_patient$exclusion_reason)),
      " patients lacked a valid non-negative analysis time.\n",
      "Panels are descriptive and populations overlap; no between-panel test was performed."
    ),
    theme = theme(
      plot.title = element_text(face = "bold", size = 15, hjust = 0.5),
      plot.subtitle = element_text(size = 11, hjust = 0.5),
      plot.caption = element_text(size = 8.5, hjust = 0)
    )
  )

print(combined_full_excluded_cif)
ggsave(
  file.path(cif_output_dir, "Supplementary_CIF_full_and_excluded.png"),
  combined_full_excluded_cif, width = 12, height = 5.5, dpi = 600, bg = "white"
)
ggsave(
  file.path(cif_output_dir, "Supplementary_CIF_full_and_excluded.pdf"),
  combined_full_excluded_cif, width = 12, height = 5.5, bg = "white"
)

stopifnot(
  full_source_n == 742L,
  full_analysis_n == 739L,
  excluded_source_n == 508L,
  excluded_analysis_n == 505L,
  nrow(prediagnosis_eskd_records) == 11L,
  nrow(full_cif_field_conflicts) == 0L
)

print(full_cif_outcome_counts)
print(full_cif_summary)
print(full_cif_disposition)
```

## 7. Berden distribution

```{r berden-figure, fig.width=7.2, fig.height=5.2}
berden_data <- RESULT_crr %>%
  transmute(
    cluster_label,
    Berden = factor(
      standardize_berden(Berden.score.on.renal.biopsy),
      levels = berden_levels
    )
  )

berden_tab <- with(berden_data, table(cluster_label, Berden))
berden_available_tab <- berden_tab[, setdiff(colnames(berden_tab), "Missing"), drop = FALSE]
if (nrow(berden_available_tab) >= 2 && ncol(berden_available_tab) >= 2) {
  set.seed(123)
  print(fisher.test(berden_available_tab, simulate.p.value = TRUE, B = 10000))
}

# Missingness is tested from patient-level rows, not from already aggregated
# category rows.
berden_missing_tab <- with(
  berden_data,
  table(cluster_label, Berden_missing = Berden == "Missing" | is.na(Berden))
)
print(berden_missing_tab)
if (all(dim(berden_missing_tab) >= 2)) print(fisher.test(berden_missing_tab))

berden_plot_data <- berden_data %>%
  count(cluster_label, Berden, .drop = FALSE, name = "n") %>%
  group_by(cluster_label) %>%
  mutate(proportion = n / sum(n)) %>%
  ungroup()

p_berden <- ggplot(
  berden_plot_data,
  aes(cluster_label, proportion, fill = Berden)
) +
  geom_col() +
  geom_text(
    aes(label = if_else(proportion > 0, scales::percent(proportion, accuracy = 1), "")),
    position = position_stack(vjust = 0.5),
    size = 4.2
  ) +
  scale_y_continuous(labels = scales::percent) +
  scale_fill_manual(values = berden_cols, drop = FALSE) +
  labs(
    title = "Berden biopsy class distribution",
    x = NULL,
    y = "Proportion",
    fill = "Berden biopsy class"
  ) +
  theme_manuscript(legend = "right") +
  theme(
    plot.title = element_text(hjust = 0.5, face = "bold", size = 15),
    legend.title = element_text(face = "bold")
  )

print(p_berden)
save_plot(p_berden, "Figure_3_Berden_distribution.png", 7.2, 5.2)
```

## 8. Supplementary tables and exported clustering data

```{r supplementary-tables}
summarise_characteristics <- function(data, groups = character()) {
  if (length(groups)) data <- data %>% group_by(across(all_of(groups)))
  data %>%
    summarise(
      N = n(),
      mean_age = round(median(ageAtDiagnosis, na.rm = TRUE), 1),
      age_q1 = round(quantile(ageAtDiagnosis, 0.25, na.rm = TRUE), 1),
      age_q3 = round(quantile(ageAtDiagnosis, 0.75, na.rm = TRUE), 1),
      female_n = sum(gender == "Female", na.rm = TRUE),
      mean_creatinine = round(median(creatininetest1, na.rm = TRUE), 0),
      creatinine_q1 = round(quantile(creatininetest1, 0.25, na.rm = TRUE), 0),
      creatinine_q3 = round(quantile(creatininetest1, 0.75, na.rm = TRUE), 0),
      mean_egfr = round(median(eGFR1, na.rm = TRUE), 0),
      egfr_q1 = round(quantile(eGFR1, 0.25, na.rm = TRUE), 0),
      egfr_q3 = round(quantile(eGFR1, 0.75, na.rm = TRUE), 0),
      MPA_n = sum(DefiniteMPA == 1, na.rm = TRUE),
      GPA_n = sum(DefiniteGPA == 1, na.rm = TRUE),
      Berden_Crescentic = sum(Berden.score.on.renal.biopsy == "Crescentic", na.rm = TRUE),
      Berden_Focal = sum(Berden.score.on.renal.biopsy == "Focal", na.rm = TRUE),
      Berden_Mixed = sum(Berden.score.on.renal.biopsy == "Mixed", na.rm = TRUE),
      Berden_Sclerotic = sum(Berden.score.on.renal.biopsy == "Sclerotic", na.rm = TRUE),
      Berden_Missing = sum(Berden.score.on.renal.biopsy == "Missing", na.rm = TRUE),
      MPO_n = sum(ancaSpec == "MPO", na.rm = TRUE),
      PR3_n = sum(ancaSpec == "PR3", na.rm = TRUE),
      Other_n = sum(ancaSpec == "Other", na.rm = TRUE),
      Ind_Cyclo = sum(inductionFinal == "Cyclo", na.rm = TRUE),
      Ind_Rituximab = sum(inductionFinal == "Rituximab", na.rm = TRUE),
      Ind_Others = sum(inductionFinal == "Other", na.rm = TRUE),
      Dialysis_dependent_n = sum(Dialysis.dependent == "Yes", na.rm = TRUE),
      Corti_Oral = sum(
        inductiontreatmentType_Oral.corticosteroids == "Oral.Steroids",
        na.rm = TRUE
      ),
      Corti_Intra = sum(
        inductiontreatmentType_.Pulsed.IV.corticosteroids == "IV.Steroids",
        na.rm = TRUE
      ),
      Plasma = sum(inductiontreatmentType_Plasma.exchange == "Checked", na.rm = TRUE),
      .groups = "drop"
    )
}

Full_cohort_unique <- Full_cohort %>%
  distinct(RKD.ID, .keep_all = TRUE) %>%
  rename(eGFR1 = egfr_calculated1)

Full_TABLE <- summarise_characteristics(Full_cohort_unique) %>%
  rename(Ind_other = Ind_Others)

Ect_data_unique <- Ect_data %>%
  distinct(RKD.ID, .keep_all = TRUE) %>%
  mutate(inductionFinal = as.character(inductionFinal))

ECT_TABLE <- bind_rows(
  Ect_data_unique,
  Ect_data_unique %>% mutate(inductionFinal = "Overall")
) %>%
  summarise_characteristics("inductionFinal") %>%
  select(-Ind_Cyclo, -Ind_Rituximab, -Ind_Others) %>%
  arrange(inductionFinal == "Overall")

RETRACE_summary_source <- bind_rows(
  RESULT_crr %>% mutate(cluster_label = as.character(cluster_label)),
  RESULT_crr %>% mutate(cluster_label = "Overall")
)

RETRACE_TABLE <- summarise_characteristics(RETRACE_summary_source, "cluster_label") %>%
  left_join(
    RETRACE_summary_source %>%
      group_by(cluster_label) %>%
      summarise(
        ESKD_n = sum(status == 1L),
        Death_competing_n = sum(status == 2L),
        .groups = "drop"
      ),
    by = "cluster_label"
  ) %>%
  arrange(cluster_label == "Overall")

RETRACE_TABLE_WIDE <- RETRACE_TABLE %>%
  pivot_longer(-cluster_label, names_to = "Characteristic", values_to = "Value") %>%
  pivot_wider(names_from = cluster_label, values_from = Value) %>%
  relocate(Overall, .after = last_col())

# Preserve the original Table 3 P-value definitions for compatibility.
test_data <- RESULT_crr %>%
  mutate(
    diagnosis_group = case_when(
      DefiniteMPA == 1 ~ "MPA",
      DefiniteGPA == 1 ~ "GPA",
      TRUE ~ NA_character_
    ),
    oral_steroid = inductiontreatmentType_Oral.corticosteroids == "Oral.Steroids",
    iv_steroid = inductiontreatmentType_.Pulsed.IV.corticosteroids == "IV.Steroids",
    plasma_exchange = inductiontreatmentType_Plasma.exchange == "Checked"
  )
g <- test_data$cluster_label

p_values <- tibble(
  Characteristic = c(
    "mean_age", "female_n", "mean_creatinine", "mean_egfr", "MPA_n",
    "Berden_Crescentic", "MPO_n", "Ind_Cyclo", "Dialysis_dependent_n",
    "Corti_Oral", "Corti_Intra", "Plasma", "ESKD_n"
  ),
  p_raw = c(
    wilcox_p(test_data$ageAtDiagnosis, g),
    fisher_p(test_data$gender, g),
    wilcox_p(test_data$creatininetest1, g),
    wilcox_p(test_data$eGFR1, g),
    fisher_p(test_data$diagnosis_group, g),
    fisher_p(test_data$Berden.score.on.renal.biopsy, g),
    fisher_p(test_data$ancaSpec, g),
    fisher_p(test_data$inductionFinal, g),
    fisher_p(test_data$Dialysis.dependent, g),
    fisher_p(test_data$oral_steroid, g),
    fisher_p(test_data$iv_steroid, g),
    fisher_p(test_data$plasma_exchange, g),
    fisher_p(test_data$status, g)
  )
) %>%
  mutate(`P value` = if_else(is.na(p_raw), "", vapply(p_raw, format_p, character(1))))

RETRACE_TABLE_WIDE <- RETRACE_TABLE_WIDE %>%
  left_join(p_values %>% select(Characteristic, `P value`), by = "Characteristic") %>%
  relocate(`P value`, .after = last_col())

save_table(Full_TABLE, "supplementary_table_1_full_cohort_egfr.csv")
save_table(ECT_TABLE, "supplementary_table_2_ECT_cohort_egfr.csv")
save_table(RETRACE_TABLE, "supplementary_table_3_RETRACE_cohort_egfr.csv")
save_table(
  RETRACE_TABLE_WIDE,
  "supplementary_table_3_RETRACE_cohort_egfr_wide_with_p_values.csv"
)

print(Full_TABLE)
print(ECT_TABLE)
print(RETRACE_TABLE_WIDE)
```

```{r cluster-centroids}
cluster_centroids_long <- Data_wide %>%
  transmute(
    RKD.ID,
    cluster_label = as.character(cluster_label),
    time_bin,
    eGFR_change = num_clean(mean)
  ) %>%
  group_by(RKD.ID, cluster_label, time_bin) %>%
  summarise(
    eGFR_change = if (all(is.na(eGFR_change))) NA_real_ else mean(eGFR_change, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  group_by(cluster_label, time_bin) %>%
  summarise(
    centroid = mean(eGFR_change, na.rm = TRUE),
    n_available = sum(!is.na(eGFR_change)),
    .groups = "drop"
  ) %>%
  mutate(centroid = if_else(is.nan(centroid), NA_real_, centroid)) %>%
  arrange(cluster_label, time_bin)

cluster_centroids_wide <- cluster_centroids_long %>%
  select(cluster_label, time_bin, centroid) %>%
  mutate(time_bin = paste0("month_bin_", time_bin)) %>%
  pivot_wider(names_from = time_bin, values_from = centroid)

save_table(cluster_centroids_long, "eGFR_cluster_centroids_long.csv")
save_table(cluster_centroids_wide, "eGFR_cluster_centroids_wide.csv")
save_table(kml_wide, "kml_input.csv")
```

