---
title: "eGFR workflow"
author: "Shen Jia"
date: "2026-07-01"
output:
  html_document:
    toc: true
    toc_depth: 2
params:
  project_root: "."
  data_path: "data/harmonized_data.csv"
  output_dir: "outputs/interactive_run"
---

```{r load-analysis-code, include=FALSE}
# kml imports rgl; use its headless device when this report is rendered without
# a graphical display.
options(rgl.useNULL = TRUE)

# The step files contain publication labels such as µ, ², and en dashes.
# Set the character locale before knitr reads and parses those files.
if (!isTRUE(l10n_info()[["UTF-8"]])) {
  for (locale_name in c("en_US.UTF-8", "C.UTF-8", "C.utf8")) {
    suppressWarnings(try(Sys.setlocale("LC_CTYPE", locale_name), silent = TRUE))
    if (isTRUE(l10n_info()[["UTF-8"]])) break
  }
}
if (!isTRUE(l10n_info()[["UTF-8"]])) {
  warning("No UTF-8 locale is available; non-ASCII labels may be escaped.")
}

project_root <- normalizePath(params$project_root, mustWork = TRUE)
knitr::opts_knit$set(root.dir = project_root)
step_files <- sort(list.files(file.path(project_root, "analysis", "steps"),
                              pattern = "[.]R$", full.names = TRUE))
read_step <- knitr::read_chunk
for (step_file in step_files) read_step(step_file)
```

Run the analysis in the order shown below. Code is loaded from `analysis/steps/`; shared functions are in `R/analysis_helpers.R`.
```{r setup, message=FALSE, warning=FALSE}
```
```{r analysis-paths}
```
## 1. Load, filter, and clean the data
Select the analysis variables and derive the full AAV cohort.

```{r load-filter-clean}
```
## 2. Derive the early renal recovery cohort (Cohort 1)

```{r cohort-derivation}
```
## 3. Early eGFR recovery analysis in Cohort 1

Model change in eGFR during the first six months after diagnosis using mixed-effects models with prespecified covariates.

```{r ect-models}
```
```{r ect-figures, fig.width=7.2, fig.height=8.5}
```
## 4. Identification of early eGFR recovery groups (Cohort 2)

```{r retrace-kml}
```
```{r cluster-selection-figure, fig.width=7.4, fig.height=4.6}
```
```{r trajectory-figure, fig.width=7.2, fig.height=5.2}
```
## 5. RETRACE outcome analysis

```{r fine-gray-models}
```
```{r retrace-cif, fig.width=7.2, fig.height=5.35}
```

## 6. Post-peak eGFR slope analysis set (Cohort 3)

T1 is the earliest occurrence of the highest available eGFR between 4 and 12
months after diagnosis. Dialysis-dependent visits are assigned an eGFR of 3
mL/min/1.73 m². Measurements from T1 through 60 months after diagnosis are
retained, including measurements after ESKD; follow-up ends at transplantation,
death, or last recorded contact. Eligibility requires T1 and at least one later
eGFR measurement on a distinct date.

```{r post-peak-slope-data}
```
```{r post-peak-slope-model-and-figures, fig.width=10.8, fig.height=10.3}
```
### Cohort 3 characteristics

```{r cohort3-characteristics-table}
```
## 7. Supplementary full and RETRACE-excluded CIFs

This analysis is descriptive: the excluded cohort is a subset of the full
cohort, so the panels must not be compared with a between-group test.

```{r supplementary-full-and-excluded-cif, fig.width=12, fig.height=5.5}
```
## 8. Berden distribution

```{r berden-figure, fig.width=7.2, fig.height=5.2}
```

## 9. Supplementary tables and exported clustering data

```{r supplementary-tables}
```
```{r cluster-centroids}
```
